### 2024-10-02 README.Rmd
Update abstract; add title and authors

Understanding the processes that control population dynamics in wild populations is a critical component of conservation and management but is frequently limited by missing demographic information, uncertainty, and temporally or spatially misaligned datasets (Schaub & Abadi 2011, Zipkin & Saunders 2017). Integrated population models (IPMs; Besbeas et al. 2002, Brooks et al. 2004) are an increasingly popular tool in ecology that can help overcome these limitations by combining disparate datasets in a single, unified analysis (Schaub & Abadi 2011). This approach can provide several advantages, such as improved precision (Schaub et al. 2007, Tavecchia 2009, Abadi et al. 2010) and the estimation of parameters that would otherwise be unidentifiable (Besbeas 2005, Schaub et al. 2007, Veran & Lebreton 2008), which is often the case for key demographic rates such as recruitment or immigration. These advances have practical benefits for resource managers that must make decisions based on limited available information. As IPMs become more widely used, however, it is critical that they are robustly evaluated. In particular, the limitations of the IPM framework are still underexplored, particularly in situations where available data are sparse, of poor quality, or when data to inform specific parameters are entirely absent.

Through a simulation analysis, we will identify monitoring programs and demographic scenarios where integrated population models may not lead to improved precision or return unbiased estimates of parameters that would otherwise be unidentifiable when certain datasets are excluded. We apply this simulation study to to a passerine bird life history. We assume all model assumptions are met and compare parameter estimation for a traditional IPM with data informing abundance (count data), survival (mark-recapture data), and reproductive output (nest monitoring data) to situations when either of the latter two datasets are omitted. In doing so, we will refine our understanding of the mechanisms underlying IPMs and can better identify when they may serve as a useful tool or might not be necessary or warrant the collection of additional data. Improving our understanding of the strengths and weaknesses of the IPM framework can provide helpful insights for the design and implementation of field survey programs so as to most efficiently make use of limited resources to inform management efforts aimed at monitoring or recovering species that are deemed ecosystem indicators or depleted and in need of conservation intervention.

# Scripts, current version

## 0 - preparing scenarios

### compute_time_calc.R
Code for approximating the total run time, given number of simulated datasets, number of parameter scenarios, computer cores, and estimated run time for each model.

### generate_scenarios.R

@@

## 1 - simulating data

### IPM_sim_2.0function.R

@@

### simulateTrajectories.R

@@

## 2 - models

### IPMinitvalues.R
Functions for putting data in the form required by the survival model.

### IPM_marray.R
Script contains 4 IPM NIMBLE models.

@@

## 3 - run models

### run_scenarios_helperFns.R

@@

### run_scenarios.R

@@

## 4 - process results

### 01_load_and_process_data.R

@@

### 02_compute_geom_means.R

@@

# Data

>>>>

# Effects of species life history and data availability on integrate population model performance

Abby E. Bratt, Caroline D. Cappello, Amelia J. DuVall, Hannah A. Sipe, Amanda J.
Warlick, Beth Gardner, Sarah J. Converse

Code provided for peer review and may change

## Abstract

Integrated population models (IPMs) are an increasingly popular tool for population modeling in ecology. By combining datasets in a unified analysis, IPMs link demographic rates with population dynamics and can thereby reduce bias, improve precision, and estimate parameters that would otherwise be unidentifiable using traditional approaches. However, the performance of IPMs in different situations remains underexplored and the conditions under which IPMs produce more precise estimates compared with those produced in singular analyses are poorly understood.

We assessed the performance of IPMs across a range of life history parameters, population trajectories, and data availability scenarios using simulated count, mark-resight, and productivity data typical of a passerine monitoring program. We examined relative bias and root mean squared error for abundance, trend in abundance, survival, and fecundity across models when all datasets were included versus when one or two datasets were omitted or when the included datasets were of poorer quality (i.e., lower detection probabilities).

Parameter estimates generally showed higher bias when the datasets that more directly informed those parameters were omitted or when the true data-generating parameters were representative of more extreme life history traits (e.g., high survival and low reproduction that characterizes large mammals or the lower survival and higher reproductive output that characterizes insects or amphibians). Higher detection probabilities alone did little to reduce bias in demographic rates. However, higher detection probabilities in count surveys were more effective at reducing uncertainty in abundance trends than using more of the available datasets.

Our results suggest that the unique context of a given monitoring program and study species will largely determine whether collecting multiple data streams versus collecting a single high-quality data stream (e.g., increasing detection probabilities or sample sizes) will most improve parameter estimates. Practitioners should carefully consider the best ways of answering the specific ecological or management questions of interest (e.g., estimating demographic rates versus detecting trends) given the species’ life history and the costs of data collection. This work provides a foundation from which to further explore the performance of IPMs and to inform the design of monitoring programs while considering the costs and benefits of data collection.

## Scripts, current version

### 0 - preparing scenarios

#### compute_time_calc.R
Code for approximating the total run time, given number of simulated datasets, number of parameter scenarios, computer cores, and estimated run time for each model.

#### generate_scenarios.R

@@

### 1 - simulating data

#### IPM_sim_2.0function.R

@@

#### simulateTrajectories.R

@@

### 2 - models

#### IPMinitvalues.R
Functions for putting data in the form required by the survival model.

#### IPM_marray.R
Script contains 4 IPM NIMBLE models.

@@

### 3 - run models

#### run_scenarios_helperFns.R

@@

#### run_scenarios.R

@@

### 4 - process results

#### 01_load_and_process_data.R

@@

#### 02_compute_geom_means.R

@@

## Data
